Multiple Comparisons in Toolmark Evidence

### 2023-11-20 LPR/index.qmd
Make comparison counts more general

@@

Let us first consider the situation where a single tool is recovered from the suspect's house that is viewed as likely to have made the wire cut. Then, a forensic examiner might take the tool and use it to cut a piece of metal, known as a blade cut, in order to record the markings of the entire cutting surface simultaneously. This process might be repeated at multiple angles in order to record the effect of different contact surfaces on the resulting striations. 
For simplicity, let us for now assume that only a single cut under one set angle was made. Let us further define the blade to be of length $\ell$ and the wire to have diameter $d$ which is fully covered with striations that are untouched by the blast.

At minimum, the examiner will have to do $\ell/d$ comparisons in order to assess whether the tool matches the wire; when a comparison algorithm is used, and both the blade and the wire are scanned at the same resolution of $r$ mm, there may be as many as $\ell/r- d/r + 1$ comparisons performed automatically in order to find the optimal alignment between the two samples[^1].

Ultimately, in order to compare tool marks in this scenario, the examiner or algorithm completes the equivalent to a database search resulting from a single piece of evidence. The dangers of such searches are well known and have been cited in forensic missteps such as the Madrid bombing case, as discussed in @li2023; @newman2007; @mayfield2006report [Pg 137].

>>>>

Let us consider the process of toolmark examination - an examiner is given a tool recovered from a suspect as well as a wire recovered from the scene. The examiner will typically take the tool and use it to cut a piece of metal, known as a blade cut, in order to record the markings of the entire cutting surface simultaneously. This process may be repeated at multiple angles in order to record the effect of different contact surfaces on the resulting striations. The resulting blade cut will be compared to the wire using either visual inspection or perhaps in the future a numerical comparison algorithm similar to those developed for bullet striations [@hare2017automatic] and cartridge cases [@zemmels2023].

The examiner will compare blade cuts from all cutting surfaces during this process. Tools used for wire cuts typically have either 2 or 4 cutting surfaces. Examiners may make multiple cuts at multiple angles; this will increase the number of available blade cuts for comparison.

For simplicity, let us for now assume that only a single cut under one controlled angle was made.

### Counting Comparisons

Let us consider a single blade cut from a single cutting surface. Define the blade cut to be of length $\ell$ and the wire to have diameter $d$; we will also assume that the wire is fully covered with striations suitable for comparison.

At minimum, there must be $\ell/d$ comparisons in order to assess whether the tool matches the wire; in this case, the comparisons are essentially independent of each other (conditional on the striation patterns) because no overlapping data is used. Of course, in this situation, we are overlooking the potential for partial matches and assuming that everything is lined up perfectly to produce this minimum number of comparisons; in practice, of course, the number of comparisons in this situation would likely be much higher.

At a maximum, both the blade and the wire are either digitally scanned at the same resolution of $r$ mm, or visually examined using a microscope with a digital view of a certain resolution which can be translated into an equivalent resolution to the digital scan. In this case, there may be as many as $\ell/r- d/r + 1$ comparisons performed in order to find the optimal alignment between the two samples[^1]. Of course, these comparisons are highly dependent, as each sequential comparison shares much of the same data with the previous and next comparisons. While it is likely possible to model this autocorrelation and compute an effective number of comparisons accounting for the autocorrelation, developing this model is outside the scope of this paper.

Thus, we will consider the two extremes: $\ell/d$ completely independent comparisons and $\ell/r - d/r + 1$ dependent comparisons. The true number of visual or numerical comparisons lies somewhere between these two values.

@@

Let us first consider the situation where a single tool is recovered from the suspect's house that is viewed as likely to have made the wire cut. Ultimately, in order to compare tool marks in this scenario, the examiner or algorithm completes the equivalent to a database search resulting from a single piece of evidence. The dangers of such searches are well known and have been cited in forensic missteps such as the Madrid bombing case, as discussed in @li2023; @newman2007; @mayfield2006report [Pg 137].
